A formal approach but practical approach in train regulation

### 2025-06-24 paper/main.tex
Ajout des règles du régulateur

%% Regulator's Program
\newcommand{\regTuple}[3]{\tuple{#1, #2, #3}}
\newcommand{\evGuard}[2]{\ensuremath{\mathtt{ev}(\trainFmt{#1}, #2)}}
\newcommand{\ev}[3]{\evGuard{#1}{#2}: #3}
\newcommand{\incr}[1]{{\mbox{\ensuremath{\mathtt{incr}(\posFmt{#1})}}}\xspace}
\newcommand{\turn}[2]{{\mbox{\ensuremath{\mathtt{turn}(#1, #2)}}}\xspace}
\newcommand{\tokens}{\ensuremath{T}}
\newcommand{\tokenOf}[1]{\ensuremath{T(\posFmt{#1})}}
\newcommand{\regulator}{\ensuremath{R}}
\newcommand{\wait}[2]{{\mbox{\ensuremath{\mathtt{wait}(\posFmt{#1}, #2)}}}\xspace}
\newcommand{\regConcat}[2]{#1 \cdot #2}

@@

%% tt style

@@

\begin{itemize}
	\item The routing program $E$ specifies how the regulator should react to events (i.e. when a train enters a block). A program is a collection of rules to apply upon each event (called an \emph{event handler}). We detail further event handlers below.
	\item The token list, denoted $\tokens$, is a function  $\tokens: \blocks \to \Nat$ associating a token level to each block. Those tokens are used to prevent multiple trains accessing the same block at the same time.
	\item The waiting set is a set $W \subseteq \blocks \times \Nat \times \trains$. A triple $\tuple{\bid{id}, n, \tid{id}} \in W$ means block $\bid{id}$ must be protected against train $\tid{id}$ entering as long as $\tokenOf{\bid{id}} < n$.
\end{itemize}

@@

Train events are handled by the regulator. Upon such events, the regulator can take actions such as turning a switch, allowing or blocking a train, etc.. The regulator is equipped with a routing program. This routing program contains, for each train, a list of handlers. For a given train, the first handler of the list is the set of actions that have to be executed on the train's first event, second handler for the second event, and so on.

@@

The syntax of each handler is given by the non-terminal {\ttfamily stmt} in Grammar~\ref{gr:regulator_program}.\todo{est-ce toujours nécessaire maintenant que les handlers sont dans un liste ?}
The program of the regulator is a set of rules, guarded by events. A rule $\ev{\tid{id}}{n}{A}$ denotes that upon entering the $n$-th block of its route, train $\tid{id}$ will trigger actions $A$. The list of actions in the handler are of four kinds: 
\begin{description}
	\item[\incr{\bid{id}}:] this updates the token $\bid{id}$ in $T$:  $T \cdot \incr{\bid{id}}  \triangleq T[\bid{id} \leftarrow T(\bid{id})+1]$
	\item [\wait{\bid{id}}{i}:] ...\todo{}
	\item [\turn{\sid{id}}{s}:] this instructs the regulator to turn the switch $\sid{id}$ into the $s$ state: $\sigma\cdot\turn{\sid{id}}{s} = \sigma[\sid{id} \leftarrow s]$
	\item [auth:] ...\todo{}
\end{description}

@@

\begin{grammar}[Regulator's rules, with $\tid{id} \in \trains$, $n \in  \Nat$, $\bid{id} \in \blocks$, $\sid{id} \in \turnouts$, $s\in \internalState$.][h][gr:regulator_program]
	\firstcasesubtil{\ttfamily stmt}{\evGuard{\tid{id}}{n} : \nonterm{A}^\star}{}
	\firstcase{A}{\incr{\bid{id}}}{}
	\otherform{\wait{\bid{id}}{i}}{}
	\otherform{\turn{\sid{id}}{s}}{}
	\otherform{auth}{}
\end{grammar}

@@

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{Start}]{
		\dirFmt{d} \neq \dirFmt{d^\prime} \and P = \trainConcat{\su{d^\prime}{\posFmt{N}}}{P^\prime}
	}{
		\hspace{3mm} \redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{P}\}}{\regulator}{\switches}{\signals}} \\
		\reduces
		\redTuple{\guardT}{\push{B}{\tuple{\tid{id}, \bid{id}}}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d^\prime}{P}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{UntilNext}] {
		\suc{\bid{id}}{d}{\switches} = \bid{id^\prime} \neq \nosuc
		\and
		\signalF{\bid{id}}{d} = \siggreen
	}{
		\redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\trainConcat{\su{d}{[\bid{id^\prime}]}}{P}}\}}{\regulator}{\switches}{\signals}} \\
		\reduces		
		\redTuple{\guardS}{\push{\bufferFmt{B}}{\tuple{\tid{id}, \bid{id^\prime}}}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id^\prime}}{d}{P}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}\todo{Pas de message dans ce cas là?}

@@

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{EndEvent}]{
		\bufhead{B} = \tuple{\tid{id}, \bid{id}}
		\and
		\head{\handlerOf{E}{\tid{id}}} = \emptyList
	}{
		\redTuple{\guardR}{\bufferFmt{B}}{\stateTuple{\trainSeq}{\tuple{E, T, W}}{\switches}{\signals}}
		\reduces
		\redTuple{\guardT}{\buftail{B}}{\stateTuple{\trainSeq}{\tuple{\popHandlerHead{E}{\tid{id}}, T, W}}{\switches}{\signals}}
	}
\end{mathpar}

@@

\begin{figure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=\textwidth]{img/spec_pc.png}
		\caption{Spec de l'ordinateur}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=\textwidth+0.5cm]{img/testElem.png}
		\caption{Vérification des chemins élémentaires}
	\end{subfigure}
	\caption{Tests sur les chemins élémentaires}
\end{figure}
\label{fig:elem-tests}

>>>>

%% Regulator's Program
\newcommand{\regTuple}[3]{\tuple{#1, #2, #3}}
\newcommand{\evGuard}[2]{\ensuremath{\mathtt{ev}(\trainFmt{#1}, #2)}}
\newcommand{\ev}[3]{\evGuard{#1}{#2}: #3}
\newcommand{\incr}[1]{{\mbox{\ensuremath{\mathtt{incr}(\posFmt{#1})}}}\xspace}
\newcommand{\turn}[2]{{\mbox{\ensuremath{\mathtt{turn}(#1, #2)}}}\xspace}
\newcommand{\auth}{{\mbox{\ensuremath{\mathtt{auth}}}}\xspace}
\newcommand{\tokens}{\ensuremath{T}}
\newcommand{\tokenOf}[1]{\ensuremath{T(\posFmt{#1})}}
\newcommand{\regulator}{\ensuremath{R}}
\newcommand{\wait}[2]{{\mbox{\ensuremath{\mathtt{wait}(\posFmt{#1}, #2)}}}\xspace}
\newcommand{\regConcat}[2]{#1 \cdot #2}

@@

%% Rule predicates
\newcommand{\requestS}[3]{\ensuremath{{\tt RequestS}(\trainFmt{#1}, \posFmt{#2}, #3)}}
\newcommand{\nextWait}[1]{\ensuremath{{\tt NextWait}(#1)}}

@@

\begin{itemize}
	\item The routing program $E$ specifies how the regulator should react to events (i.e. when a train enters a block). A program is a collection of rules to apply upon each event (called an \emph{event handler}). We detail further event handlers below.
	\item The token list, denoted $\tokens$, is a function  $\tokens: \blocks \to \Nat$ associating a token level to each block. Those tokens are used to prevent multiple trains accessing the same block at the same time.
	\item The waiting set is a set $W \subseteq \blocks \times \blocks \times \Nat \times \trains$. A triple $\tuple{\bid{src, }\bid{dst}, n, \tid{id}} \in W$ means that train \tid{id} is currently waiting in the block \bid{src} for $\tokenOf{\bid{id}}$ to be increased up to $n$.
\end{itemize}

@@

Train events are handled by the regulator. Upon such events, the regulator can take actions such as turning a switch, allowing or blocking a train, etc.. The regulator is equipped with a routing program. This routing program contains, for each train, a list of handlers\todo{Chaque handler associe une position et une liste d'ordres.}. For a given train, the first handler of the list is the set of actions that have to be executed on the train's first event, second handler for the second event, and so on.

@@

The syntax of each handler is given by the non-terminal {\ttfamily stmt} in Grammar~\ref{gr:regulator_program}.\todo{est-ce toujours nécessaire maintenant que les handlers sont dans un liste ?}
The program of the regulator is a set of rules, guarded by events. A rule $\ev{\tid{id}}{n}{A}$ denotes that upon entering the $n$-th block of its route, train $\tid{id}$ will trigger actions $A$. The list of actions in the handler are of four kinds: 
\begin{description}
	\item[\incr{\bid{id}}:] this updates the token $\bid{id}$ in $T$:  $T \cdot \incr{\bid{id}}  \triangleq T[\bid{id} \leftarrow T(\bid{id})+1]$
	\item [\wait{\bid{id}}{i}:] ...\todo{}
	\item [\turn{\sid{id}}{s}:] this instructs the regulator to turn the switch $\sid{id}$ into the $s$ state: $\sigma\cdot\turn{\sid{id}}{s} = \sigma[\sid{id} \leftarrow s]$
	\item [\auth:] ...\todo{}
\end{description}

@@

\begin{grammar}[Regulator's rules, with $\tid{id} \in \trains$, $n \in  \Nat$, $\bid{id} \in \blocks$, $\sid{id} \in \turnouts$, $s\in \internalState$.][h][gr:regulator_program]
	\firstcasesubtil{\ttfamily stmt}{\evGuard{\tid{id}}{n} : \nonterm{A}^\star}{}
	\firstcase{A}{\incr{\bid{id}}}{}
	\otherform{\wait{\bid{id}}{i}}{}
	\otherform{\turn{\sid{id}}{s}}{}
	\otherform{\auth}{}
\end{grammar}

@@

Similarly, the regulator communicates with signals with a \emph{signal buffer} $B_s$, which is a list of tuple $\tuple{\bid{id}, c}$, indicating that both signals of block $\bid{id}$ should be turned to color $c$. \todo{mise en page}

@@

\todo{Enlever les bid dans le buffer des trains}
\begin{mathpar}
	\inferrule*[left=\ruleDef{Start}]{
		\dirFmt{d} \neq \dirFmt{d^\prime} \and P = \trainConcat{\su{d^\prime}{\posFmt{N}}}{P^\prime}
	}{
		\hspace{3mm} \redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{P}\}}{\regulator}{\switches}{\signals}} \\
		\reduces
		\redTuple{\guardT}{\push{B}{\tuple{\tid{id}, \bid{id}}}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d^\prime}{P}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{UntilNext}] {
		\suc{\bid{id}}{d}{\switches} = \bid{id^\prime} \neq \nosuc
		\and
		\signalF{\bid{id}}{d} = \siggreen
	}{
		\redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\trainConcat{\su{d}{[\bid{id^\prime}]}}{P}}\}}{\regulator}{\switches}{\signals}} \\
		\reduces		
		\redTuple{\guardS}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id^\prime}}{d}{P}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}
Notice that, in \ruleRef{UntilNext}, no message is added to the buffer.
\todo{@Lucas, expliquer pourquoi c'est pas nécessaire.}

@@

Of course, trains can conflict. For instance, if two trains have to use the same track, one has to go first and the other has to wait until the track is free. To schedule trains, we use a token system. Each block is assigned a token, which is an integer value. Trains can \emph{wait} until a token reaches a given value, and \emph{increment} a token after leaving a block. We introduce two pairs of rules for incrementing and waiting. The first pair deals with the case where the train waiting arrives before the token has the correct value, and the second pair deals with the (easier) case where the train releases the track (increments the associated token) before the next train arrives (and therefore does not need to wait).

To execute a \wait{\bid{critical}}{n} order, the regulator stores the train id in its waiting set, together with the current block of the train (\bid{src}), the block it is waiting for (\bid{critical}), and the value the token associated with the target block has to reach to release the train ($n$).
\begin{mathpar}
	\inferrule*[left=\ruleDef{WaitBefore}] {
		\bufhead{B} = \tuple{\tid{id}, \bid{id}}
		\and
		\head{\handlerOf{E}{\tid{id}}} = \regConcat{\wait{\bid{critical}}{n}}{A^\prime}
		\\
		\tokenOf{\bid{critical}} \neq n
	}{
		\redTuple{\guardR}{\bufferFmt{B}}{\stateTuple{\trainSeq}{\tuple{E, T, W}}{\switches}{\signals}}
		\reduces
		\redTuple{\guardR}{\bufferFmt{B}}{\stateTuple{\trainSeq}{\tuple{\substHandlerHead{E}{\tid{id}}{A^\prime}, T, W\cup\tuple{\bid{src}, \bid{critical}, n, \tid{id}}}}{\switches}{\signals}}
	}
\end{mathpar}

Then, when the releasing train frees the critical block, the token is incremented, and the train waiting for the new value of the token is released\footnote{By construction, nothing prevents multiple trains to be waiting for the same value, thus resulting in a faulty system. However, our model checker ensures programs (including event handlers) are correct, thus detecting such case.} and updates the signals to let the waiting train go.

\begin{mathpar}
	\inferrule*[left=\ruleDef{IncrAfter}] {
		\bufhead{B} = \tuple{\tid{id}, \bid{id}}
		\and
		\head{\handlerOf{E}{\tid{id}}} = \regConcat{\incr{\bid{critical}}}{A^\prime}
		\\
		\exists \bid{src}, \tid{waiting} \text{ s.t. } \tuple{\bid{src}, \bid{critical}, \tokenOf{\bid{critical}} + 1, \tid{waiting}} \in W
		\\
		\nextWait{\handlerOf{E}{\tid{waiting}}} = \bid{next\_stop}
	}{
		\redTuple{\guardR}{\bufferFmt{B}}{\stateTuple{\trainSeq}{\tuple{E, T, W}}{\switches}{\signals}}
		\\\reduces
		\redTuple{\guardR}{\bufferFmt{B}}{\stateTuple{\trainSeq}{\tuple{\substHandlerHead{E}{\tid{id}}{A^\prime}, T[\bid{critical} \leftarrow T(\bid{critical})+1], W\setminus\{\tuple{\bid{src}, \bid{critical}, \tokenOf{\bid{critical} + 1}, \tid{waiting}}\}}}{\switches}{\signals}}
	}
\end{mathpar}
\todo{ajouter $\tuple{\bid{next\_stop, \sigred}}, \tuple{\bid{src}, \siggreen}$}
\todo{expliquer \nextWait{E}}

On the other hand, if the releasing train arrives before the waiting train, the increase rule does not need to modify the waiting set (as the waiting train is not yet in it), nor the signals\footnote{The upcoming red signal of the waiting train is dealt with in the dual \ruleRef{WaitAfter} rule.}. It only needs to increase the token. This case is detected by looking at the waiting set, when it contains no waiting train with the new value.

\begin{mathpar}
	\inferrule*[left=\ruleDef{IncrBefore}] {
		\bufhead{B} = \tuple{\tid{id}, \bid{id}}
		\and
		\head{\handlerOf{E}{\tid{id}}} = \regConcat{\incr{\bid{critical}}}{A^\prime}
		\\
		\neg\exists \bid{src}, \tid{waiting} \text{ s.t. } \tuple{\bid{src}, \bid{critical}, \tokenOf{\bid{critical}} + 1, \tid{waiting}} \in W
	}{
		\redTuple{\guardR}{\bufferFmt{B}}{\stateTuple{\trainSeq}{\tuple{E, T, W}}{\switches}{\signals}}
		\\\reduces
		\redTuple{\guardR}{\bufferFmt{B}}{\stateTuple{\trainSeq}{\tuple{\substHandlerHead{E}{\tid{id}}{A^\prime}, T[\bid{critical} \leftarrow T(\bid{critical})+1], W}}{\switches}{\signals}}
	}
\end{mathpar}

Then, when the waiting train arrives, instead of stopping the train, the regulator can update the signal and let it continue until its next waiting point. This case is identified when the token already has the value the train expects\footnote{Notice that if the train misses its slot, e.g. if an incorrect program increases twice the token, the train will not continue, as another train, waiting for the twice-increased value, might be on the tracks.}.
\begin{mathpar}
	\inferrule*[left=\ruleDef{WaitAfter}] {
		\bufhead{B} = \tuple{\tid{waiting}, \bid{id}}
		\and
		\tokenOf{\bid{critical}} = n
		\\
		\nextWait{\handlerOf{E}{\tid{waiting}}} = \bid{next\_stop}
		\and
		\head{\handlerOf{E}{\tid{id}}} = \regConcat{\wait{\bid{critical}}{n}}{A^\prime}
	}{
		\redTuple{\guardR}{\bufferFmt{B}}{\stateTuple{\trainSeq}{\tuple{E, T, W}}{\switches}{\signals}}
		\reduces
		\redTuple{\guardR}{\bufferFmt{B}}{\stateTuple{\trainSeq}{\tuple{\substHandlerHead{E}{\tid{id}}{A^\prime}, T, W}}{\switches}{\signals}}
	}
\end{mathpar}
\todo{M-à-J des feux}

\begin{mathpar}
	\inferrule*[left=\ruleDef{Auth}] {
		\bufhead{E[\tid{id}]}.pos = \bid{id}
		\and
		\bufhead{B} = \tuple{\tid{id}, \bid{id}}
		\\
		\nextWait{\buftail{E[\tid{id}]}} = \bid{wait}
		\and
		\head{\handlerOf{E}{\tid{id}}} = \regConcat{\auth}{A^\prime}
	}{
			\redTuple{\guardR}{\bufferFmt{B}}{\stateTuple{\trainSeq}{\tuple{E, T, W}}{\switches}{\signals}}
			\reduces
			\redTuple{\guardR}{\bufferFmt{B}}{\stateTuple{\trainSeq}{\tuple{\substHandlerHead{E}{\tid{id}}{A^\prime}, T, W}}{\switches}{\signals}}
	}
\end{mathpar}\todo{ajouter $\tuple{\bid{wait}, \sigred}, \tuple{\bid{id}, \siggreen}$ dans le buffer des signaux}
where \nextWait{E} returns the next position (block id) where the train has to was, using the event handlers of the train in question. Notice that, in our case, we consider the tail of the event handlers of the train, since it is already waiting and we want the next waiting location. \todo{rephrase}

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{EndEvent}]{
		\bufhead{B} = \tuple{\tid{id}, \bid{id}}
		\and
		\head{\handlerOf{E}{\tid{id}}} = \emptyList
	}{
		\redTuple{\guardR}{\bufferFmt{B}}{\stateTuple{\trainSeq}{\tuple{E, T, W}}{\switches}{\signals}}
		\reduces
		\redTuple{\guardS}{\buftail{B}}{\stateTuple{\trainSeq}{\tuple{\popHandlerHead{E}{\tid{id}}, T, W}}{\switches}{\signals}}
	}
\end{mathpar}

@@

\begin{figure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		%\includegraphics[width=\textwidth]{img/spec_pc.png}
		\caption{Spec de l'ordinateur}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		%\includegraphics[width=\textwidth+0.5cm]{img/testElem.png}
		\caption{Vérification des chemins élémentaires}
	\end{subfigure}
	\caption{Tests sur les chemins élémentaires}
\end{figure}
\label{fig:elem-tests}
